---
title: "PS5 Template"
author: "Jacob Rodriguez"
date: "2/19/2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

# Disclaimer: this serves as a template for you to compose your answer for PS4. It is 100% hardcoded, so error may exists. Please refer to the official spec for a full list of questions and requirements.

# Instructions

This is a problem set about rmarkdown and plotting (using ggplot). Unlike the previous problem sets, this one does not give you a ready-made GH repo with a code file--it is now your task to create a repo and include your rmarkdown file in there.\

You should answer the questions below in that file, knit it, and submit both the compiled html and link to your repo on canvas.\

-   This problem sets asks you to write extensively when commenting your results. Please write clearly! Answer questions in a way that if the code chunks are hidden then the result is still readable!\
    -   All substantial questions need explanations. You do not have to explain the simple things like "how many rows are there in data", but if you make a plot of life expectancy, then you should explain what does the plot tell you.\
    -   Write explanations as markdown and use the styles like bold and italic as appropriate.\
-   Do not print too much results. It is all well to print a few lines of data for evaluation/demonstration purposes. But do not print dozens (or thousands!) of lines--no one bothers to look at that many numbers. You will lose points for annoying others (here your graders, but later potentially your boss).\
-   Do not make code lines too long. 80-100 characters is a good choice. Your grader may not be able to follow all the code if the line is too long--most of us are using small laptop screens! (And again--you want to keep your graders happy!)\

# Gapminder data

We use gapminder dataset, downloaded from [**https://www.gapminder.org/data/**](https://www.gapminder.org/data/), however, the data structure there is quire complex, please use the dataset provided on canvas (in files/data).\
The variables are:\
**name** country name\
**iso3** 3-letter country code\
**iso2** 2-letter country code\
**region** broad geographic region\
**sub-region** more precise region\
**intermediate-region**\
**time** year\
**totalPopulation** total population\
**GDP_PC** GDP per capita (constant 2010 US\$)\
**accessElectricity** Access to electricity (% of population)\
**agriculturalLand** Agricultural land (sq. km)\
**agricultureTractors** Agricultural machinery, tractors (count)\
**cerealProduction** Cereal production (metric tons)\
**feritilizerHa** Fertilizer consumption (kilograms per hectare of arable land)\
**fertilityRate** total fertility rate (births per woman)\
**lifeExpectancy** Life expectancy at birth, total (years)\
**childMortality** Mortality rate, under-5 (per 1,000 live births)\
**youthFemaleLiteracy** Literacy rate, youth female (% of females ages 15-24)\
**youthMaleLiteracy** Literacy rate, youth male (% of males ages 15-24)\
**adultLiteracy** Literacy rate, adult total (% of people ages 15 and above)\
**co2** CO2 emissions (kt)\
**greenhouseGases** Total greenhouse gas emissions (kt of CO2 equivalent)\
**co2_PC** CO2 emissions (metric tons per capita)\
**pm2.5_35** PM2.5 pollution, population exposed to levels exceeding WHO Interim Target-1 value 36ug/m3\
**battleDeaths** Battle-related deaths (number of people)\

# 1 Load and check data (5pt)

You first task is to do a very simple data check:

1.  (1pt) For solving the problems, and answering the questions, create a new rmarkdown document with an appropriate title. See [**https://faculty.washington.edu/otoomet/info201-book/r-markdown.html#r-markdown-rstudio-creating**](https://faculty.washington.edu/otoomet/info201-book/r-markdown.html#r-markdown-rstudio-creating).

```{r message = FALSE}
library(tidyverse)
getwd()
```

2.  (2pt) Load data. How many rows/columns do we have?\

```{r message = FALSE}
gapmind <- read_delim("gapminder.csv.bz2")
ncol(gapmind) ##Number of columns which is 25
nrow(gapmind) ##Number of rows which is 13055
```

3.  (2pt) Print a small sample of data. Does it look OK?

```{r}
head(gapmind)
```

The data looks good but the first couple of columns are missing values and it seemed as though it was not loaded correctly but it is correct. ------------------------------------------------------------------------

# 2 Descriptive statistics (15pt)

1.  (3pt) How many countries are there in the dataset? Analyze all three: *iso3*, *iso2* and *name*.

```{r}
gapmind %>% 
  summarise(names = n_distinct(name),
            ISO2 = n_distinct(iso2),
            ISO3 = n_distinct(iso3))

```

2.  If you did this correctly, you saw that there are more iso-2 codes than names, and there are even more *iso3*-codes. What is going on? Can you find it out?

    (a) (5pt) Find how many names are there for each iso-2 code. Are there any iso-2 codes that correspond to more than one name? What are these countries?
```{r}

names <- gapmind %>%
  filter(iso2 %in% gapmind$iso2) %>%
  distinct(iso2, name)
names

```

    (b) (5pt) Now repeat the same for name and iso3-code. Are there country names that have more than one iso3-code? What are these countries? Hint: two of these entitites are *CHANISL* and *NLD CURACAO*.
```{r}
how_many <- gapmind %>%
  filter(iso3 %in% gapmind$iso3) %>%
  distinct(iso3, name)
how_many
```

3.  (2pt) What is the minimum and maximum year in these data?
```{r}
# Find maximum and minimum years
max_year <- max(gapmind$time, na.rm = TRUE)
min_year <- min(gapmind$time, na.rm = TRUE)
max_year
min_year
```

------------------------------------------------------------------------

# 3 CO2 emissions (30pt)

Next, let's analyze CO2 emissions.

1.  (2pt) How many missing co2 emissions are there for each year? Analyze both missing *CO2* and *co2_PC*. Which years have most missing data?\
```{r}
##finds the missing co2 and cos_PC values for each year
missing_co2 <- table(is.na(gapmind$co2), gapmind$time)
missing_co2
missing_co2_PC <- table(is.na(gapmind$co2_PC), gapmind$time)
missing_co2_PC
##fnds the years with the most missing data
co2_missing_rows <- which(is.na(gapmind$co2))
co2_missing_years <- table(gapmind$time[co2_missing_rows])
co2_missing_years <- sort(co2_missing_years, decreasing = TRUE)
head(co2_missing_years)
```

2.  (5pt) Make a plot of total CO2 emissions over time for the U.S, China, and India. Add a few more countries of your choice. Explain what do you see.\
```{r}
countries <- c("United States of America", "China", "India", "Mexico", "Spain")
sub_data <- gapmind[gapmind$name %in% countries, ]
ggplot(sub_data, aes(x = time, y = co2, color = name)) +
  geom_line() +
  labs(x = "Year", y = "CO2 Emissions", color = "Country") +
  ggtitle("CO2 Emissions Over Time")
```
What we see with the graph is China's Co2 emissions have grown exponentially 
more quickly than other countries like Mexico and Spain, with the US being a close contender.

3.  (5pt) Now let's analyze the CO2 emissions per capita (*co2_PC*). Make a similar plot of the same countries. What does this figure suggest?\
```{r}
countries <- c("United States of America", "China", "India", "Mexico", "Spain")
subgap <- gapmind[gapmind$name %in% countries, ]
ggplot(subgap, aes(x = time, y = co2_PC, color = name)) +
  geom_line() +
  labs(x = "Year", y = "CO2 Emissions per Capita", color = "Country") +
  ggtitle("CO2 Emissions per Capita Over Time")
```
The figure suggests that per capita emissions in the US is much higher than in China as the previous graph suggests.

4.  (6pt) Compute average CO2 emissions per capita across the continents (assume *region* is the same as continent). Comment what do you see.\
    Note: just compute averages over countries and ignore the fact that countries are of different size.\
    Hint: Americas 2016 should be 4.80.
```{r}
gapmind %>% 
  filter(time == 2016) %>% 
  group_by(region) %>% 
  summarize(avg_co2 = mean(co2_PC, na.rm = TRUE))
```

5.  (7pt) Make a barplot where you show the previous results--average CO2 emissions per capita across continents in 1960 and 2016.
    Hint: it should look something along these lines(**Please refer to the spec for the graph**)\
```{r}
co2_per_year <- gapmind %>% 
  filter(time == 1960 | time == 2016, !is.na(co2_PC), !is.na(region)) %>% 
  group_by(time, region) %>% 
  summarize(avg_co2 = mean(co2_PC, na.rm = TRUE))
ggplot(co2_per_year, aes(x = region, y = avg_co2, fill = as.factor(time))) + 
  geom_col(position = "dodge") +
  labs(title = "Average CO2 Emissions Per Capita Across Continents", x = "Continent", y = "CO2 Emissions Per Capita") +
  scale_fill_discrete(name = "Year")

```

6.  Which countries are the three largest, and three smallest CO2 emitters (in terms of CO2 per capita) in 2019 for each continent? (Assume *region* is continent).\ 
```{r}
gapmind_2016 <- gapmind %>% 
  filter(time == 2016)
top_3 <- gapmind_2016 %>% 
  group_by(region) %>% 
  top_n(3, co2_PC) %>% 
  arrange(region, desc(co2_PC))
bottom_3 <- gapmind_2016 %>% 
  group_by(region) %>% 
  top_n(-3, co2_PC) %>% 
  arrange(region, co2_PC)

print(top_3)
print(bottom_3)
      
```
The data from 2017-2019 does not work for this section so I changed it to 2016
------------------------------------------------------------------------

# 4 GDP per capita (50pt)

Let's look at GDP per capita (*GDP_PC*)

1.  (8pt) Make a scatterplot of GDP per capita versus life expectancy by country, using data for 1960. Make the point size dependent on the country size, and color those according to the continent. Feel free to adjust the plot in other ways to make it better
    Comment what do you see there
```{r}
gapmind_1960 <- gapmind %>% filter(time == 1960,!is.na(co2_PC), !is.na(region))


ggplot(gapmind_1960, aes(x = GDP_PC, y = lifeExpectancy, size = totalPopulation, color = region)) +
  geom_point() +
  scale_x_log10() +
  labs(x = "GDP per capita", y = "Life expectancy", size = "Country size", color = "Continent")

```
What we see here is that the life expectancy goes up the with higher GDP and the populations are different according to this and this is seen with sorther life expectancies for countries in Asia and Africa during the 1960s.

2.  (4pt) Make a similar plot, but this time use 2019 data only.\
```{r}
gapmind_2019 <- gapmind %>% filter(time == 2019)


ggplot(gapmind_2019, aes(x = GDP_PC, y = lifeExpectancy, size = totalPopulation, color = region)) +
  geom_point() +
  scale_x_log10() +
  labs(x = "GDP per capita", y = "Life expectancy", size = "Country size", color = "Continent")

```

3.  (6pt) Compare these two plots and comment what do you see. How has world developed through the last 60 years?\

What we can see with these two plots is that over the population of countries in Asia has grown and life expectancy has gone up while for countries in Africa has remained similar despite 60 years. For the Americqs population and life expectancy has improved much more quickly and Oceania has seen an improvement in life expectancy but not as much growth. 

4.  (6pt) Compute the average life expectancy for each continent in 1960 and 2019. Do the results fit with what do you see on the figures?\
    Note: here as *average*I mean just average over countries, ignore the fact that countries are of different size.\
```{r}
gapmind_1960 %>%
  group_by(region) %>%
  summarize(Life_exp_1960 = mean(lifeExpectancy, na.rm = TRUE))

gapmind_2019 %>%
  group_by(region) %>%
  summarize(Life_exp_2019 = mean(lifeExpectancy, na.rm = TRUE))
```

5.  (8pt) Compute the average LE growth from 1960-2019 across the continents. Show the results in the order of growth. Explain what do you see.\
    Hint: these data (data in long form) is not the simplest to compute growth. But you may want to check out the lag() function. And do not forget to group data by continent when using lag(), otherwise your results will be messed up! See [**https://faculty.washington.edu/otoomet/info201-book/dplyr.html#dplyr-helpers-compute**](https://faculty.washington.edu/otoomet/info201-book/dplyr.html#dplyr-helpers-compute).\
```{r}

avg_growth <- gapmind_1960 %>% 
  left_join(gapmind_2019, by = c("name", "region")) %>% 
  mutate(le_growth = lifeExpectancy.y - lifeExpectancy.x) %>% 
  group_by(region) %>% 
  summarize(avg_le_growth = mean(le_growth, na.rm = TRUE)) %>% 
  arrange(avg_le_growth, increasing = TRUE)

# Show the results in the order of growth
avg_growth


```
What we see here is that the different regions life expectancy went up and in regions such as Asia and Africa it going up by 23 years while in Europe only by 11 years.

6.  (6pt) Show the histogram of GDP per capita for years of 1960 and 2019. Try to put both histograms on the same graph, see how well you can do it!\
```{r}
ggplot() +
  geom_histogram(data = gapmind_1960, aes(x = GDP_PC, fill = "1960"), alpha = 0.5) +
  geom_histogram(data = gapmind_2019, aes(x = GDP_PC, fill = "2019"), alpha = 0.5) +
  # Set axis labels and legend title
  labs(x = "GDP per capita", fill = "Year")

```


7.  (6pt) What was the ranking of US in terms of life expectancy in 1960 and in 2019? (When counting from top.)\
    Hint: check out the function rank()!\
    Hint2: 17 for 1960.\
```{r}

us_rank <- gapmind %>% 
  filter(time == 1960 | time == 2019) %>% 
  group_by(time) %>% 
  mutate(lifeExp_rank = rank(lifeExpectancy))


us_rank_1960 <- us_rank %>% 
  filter(time == 1960 & name == "United States of America") %>% 
  pull(lifeExp_rank)

us_rank_2019 <- us_rank %>% 
  filter(time == 2019 & name == "United States of America") %>% 
  pull(lifeExp_rank)

us_rank_1960
us_rank_2019

```
I am not sure what I did incorrect but the rank is showing up as this rather than 17 but if this is the case than we would see that the rank fell 135 places

8.  (6pt) If you did this correctly, then you noticed that US ranking has been falling quite a bit. But we also have more countries in 2019--what about the relative rank divided by the corresponding number of countries that have LE data in the corresponding year?\
    Hint: 0.0904 for 1960.\
```{r}
gapmind %>%
  filter(time == 1960 | time == 2019) %>%
  group_by(time) %>%
  mutate(lifeExp_rank = dense_rank(desc(lifeExpectancy))) %>%
  group_by(time) %>%
  mutate(rel_lifeExp_rank = lifeExp_rank / n_distinct(name[!is.na(lifeExpectancy)])) %>%
  filter(name == "United States of America") %>%
  select(time, lifeExpectancy, lifeExp_rank, rel_lifeExp_rank)

```
    
------------------------------------------------------------------------

Finally tell us how many hours did you spend on this PS.
I spent about 10-11 hours on this.
